---
title: "R for Data Science"
subtitle: 'Iteration with purrr'
author: "Qi Guo"
date: "August 19, 2019"
output:
  ioslides_presentation: 
    smaller: true
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, tidy = FALSE, collapse = TRUE, cache = TRUE)
library(tidyverse)
library(modelr)
options(na.action = na.warn)
```

# Part I: Introduction

## Reduce Duplication

* In `Functions`, if we copied and pasted a block of code more than twice, it's recommend to write a function to replace them, actually the loops is a kind of function, another tool for reducing duplication. For example:

```{r}
df<- tibble(
  a =  rnorm(10),
  b =  rnorm(10),
  c =  rnorm(10),
  d =  rnorm(10)
)
```

* To find a median of each column.
```{r}
median(df$a)
median(df$b)
median(df$c)
median(df$d)
```


# Part II: For Loops

## For loops

* A common template for loop:
```{r}
output <- vector("double", ncol(df))    # 1. output
for (i in seq_along(df))  {             # 2. sequence
  output[[i]] <- median(df[[i]])        # 3. body
}
output
```

* In `Functions`, that question in `Exercises` can write in:
```{r}
rescale01 <- function(x){
rng<- range(x, na.rm = TRUE)
(x - rng[1]) / (rng[2] - rng[1])
}

for (i in seq_along(df)){
  df[[i]] <- rescale01(df[[i]])
}
```

## Exercise
* 1.Write a for loop to compute different types of functions for a dataset. 
```{r}
looper <- function(the_type, object, fun) {
  empty_vec = vector(the_type, length(object))
  for (every_column in seq_along(object)) {
    empty_vec[every_column] <- fun(object[[every_column]])  
  }
  empty_vec
}
```
* So if we would like to compute the mean of every column in `mtcars`
```{r}
looper("numeric", mtcars, mean)
```
* determine the type of each column in `nycflights13::flights`
```{r, warning=FALSE}
looper("character", nycflights13::flights, class)
```

## Exercise
* \ 2. Write a function that prints the mean of each numeric column in a data frame, along with its name. For examplem, `show_mean(iris)` would print:
```{r, eval= FALSE}
show_mean(iris)
#> Sepal.Length: 5.84
#> Sepal.Width: 3.06
#> Petal.Length: 3.76
#> Petal.Width: 1.20
```

```{r}
show_means <- function(x) {
  the_class <- vector("logical", length(x))
  for (i in seq_along(x)) the_class[[i]] <- is.numeric(x[[i]])
  x <- x[the_class]
  for (i in seq_along(x)) {
    cat(paste0(names(x)[i], ": ", round(mean(x[[i]]), 2)), fill = TRUE) 
  }
}
show_means(iris)
```

# Part III: The Map Functions

## The suffix of map function
#### The type of the vector is determined by the suffix to the map function:
* `map()` makes a list.
* `map_lgl()` makes a logical vector.
* `map_int()` makes a integer vector.
* `map_dbl()` makes a double  vector.
* `map_chr()` makes a character vector.
```{r}
map_dbl(df, mean)
```
 or
```{r}
df %>% map_dbl(mean)
```

## Shortcuts
* There are a few shortcuts that use with `.f` in order to save a little typing. For example:
```{r}
models <- mtcars %>%
  split(.$cyl) %>%
  map(function(df) lm(mpg ~ wt, data = df))
```
* The syntax for creating an anonymous function in R is quite verbose so `purrr` provides a convenient shortcut:
```{r}
models <- mtcars %>%
  split(.$cyl) %>%
  map(~lm(mpg~ wt, data = .))
```

## Exercise 

* \ 1.Now use the `map()` to determine the type of each column in `nycflights13:flights`.
```{r, echo=FALSE}
map(nycflights13::flights, class)
```


## Exercise 
* \ 2.Write a function compute the number of unique values in each column of `iris`.
```{r}
map(iris, ~length(unique(.)))
map(iris, function(x) sum(table(unique(x))))
```

## Mapping over Multiple Arguments
* When the input is not a single one, `map2()` and `pmap()` are recommended to use. For example:
```{r}
mu <- list(5, 10, -3)
mu %>%
  map(rnorm, n = 5) %>%
  str()
```
* If we want to vary the standard deviation.
```{r}
sigma <- list(1, 5, 10)
seq_along(mu) %>%
  map(~rnorm(5, mu[[.]], sigma[[.]])) %>%
  str()
```

## `map2()`
* Use the `map2()`.
```{r}
map2(mu, sigma, rnorm, n = 5) %>% str()
```

* Actually `map2()` is just a wrapper around a for loop:
```{r, eval=FALSE}
map2 <- function(x, y, f, ...){
  out <- vector("list", length(x))
  for (i in seq_along(x)){
    out[[i]] <- f(x[[i]], y[[i]], ...)
  }
  out
}
```


## `pmap()`
* When we take a list of arguments, like vary the mean, standard deviation, and number of samples.
```{r}
n <- list(1,3,5)
args1 <- list(n, mu, sigma)
args1 %>%
  pmap(rnorm) %>% 
  str()
```

## Invoking Different Functions
* If we might also vary the function itself, use `invoke_map()` function.
```{r}
f <- c("runif", "rnorm", "rpois")
param <- list(
  list(min = -1, max = 1),
  list(sd = 5),
  list(lambda = 10)
)
invoke_map(f, param, n=5) %>% str()
```


# Part IV: Brainstorm
## Write a function to keep the numeric columns in a dataset.
```{r}
df <- tribble(
~x, ~y, ~z,
"a", 2, 3.6,
"b", 1, 8.5
)
df
```

```{r}
the_class <- vector("logical", length(df))
for (i in seq_along(df)) the_class[[i]] <- is.numeric(df[[i]])
   df1 <- df[the_class]
   df1
```


## Write a function to keep the numeric columns in a dataset.
```{r}
is_num <- sapply(df, is.numeric)
df2 <- df[, is_num]
df2
```

```{r}
# `detect` function
df3<- keep(df, is.numeric)
df3
```
