---
title: "R for Data Science"
subtitle: 'Data Transformation with dplyr'
author: "Qi Guo"
date: "April 10, 2019"
output:
  ioslides_presentation: 
    smaller: true
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, prompt = TRUE, tidy = FALSE, collapse = TRUE, cache = TRUE)
library(tidyverse)
library(nycflights13)
library(modelr)
options(na.action = na.warn)
```

# Part 1: Introduction

## Transformation

###Often we need to create some new variables or summarises, or maybe rename the variables or reorder the observations in order to make the data a little easier to work with


## Package

###Here we focus on the `dplyr` package, another core member of the `tidyverse`, and we'll illustrate the key ideas using data from the `nycflights13` package, and use `ggplot2` to help us understand the data.
```{r, eval=FALSE}
library(nycflights13)
library(tidyverse)
```

## Data
```{r}
flights %>% print(width = Inf)
```

## Functions
### We introduce kinds of functions to transform data, like `filter`, `arrange`, `summarise`... in next several parts

# Part II: filter

## `filter`

### `filter` allows you to subset observations based on their values.
### For example: select all flights on January 1st
```{r}
filter(flights, month==1, day==1)
```

## Comparison
### Computers use finite precision arithmetic, so every number we see is an approximation. Instead of relying on `==`, use `near`:
```{r}
sqrt(2)^2==2
near(sqrt(2)^2,2)
1/49 *49==1
near(1/49 *49,1)
```

## x %in% y
### Select every row where x is one of the values in y
### For example: select all flights that departed in November or December
```{r}
filter(flights, month %in% c(11,12))
```

## Different logical operators
### `&` is "and", `|` is "or", and `!` is "not"
### For example: select flights that weren't delayed by more than two hours
```{r}
f1<-filter(flights,!(arr_delay > 120|dep_delay>120))
# or
f2<-filter(flights, arr_delay <= 120, dep_delay <= 120)
identical(f1,f2)
```


## Missing Values
### `NA` represents an unknown value so missing values are "contagious"; almost any operation involving an unknown value will also be unknown
```{r}
NA > 5
10 == NA
NA == NA
```
### Check if a value is missing, use `is.na`
```{r}
x <- NA
is.na(x)
```

## `NA` in `filter`
### `filter` only includes rows where the condition is `TRUE`; it exculdes both `NA` and `FALSE` values, ask for them explicitly if we want to preserve missing values
```{r}
df<- tibble(x=c(1,NA,3))
filter(df,x>1)
filter(df,is.na(x) | x>1)
```

## `between`
###filter the observations between two values
### For example, find all flights departed between midnight and 6 a.m(inclusive)
```{r}
f3<-filter(flights, !(dep_time >= 2400| dep_time <=600))
f4<-filter(flights, between(dep_time, 601, 2359))
identical(f3,f4)
```

## Exercise 1

###1. Finds all flights departed in summer(July, August, and September)

\


###2.How many flights have a missing `dep_time`? What other variables are missing?


## Solution

###Q1. Finds all flights departed in summer(July, August, and September)
```{r}
filter(flights, month %in% c(7:9))
```

##Q2 How many flights have a missing `dep_time`?
```{r}
sum(is.na(flights$dep_time))
```
```{r}
#or
flights %>% select(dep_time) %>%
  is.na() %>% sum()
#or
flights %>% select(dep_time) %>%
  summarise(missing = sum(is.na(dep_time)))
```

## What other variables are missing?

###`map_dbl` function:transform their input by applying a function to each element and returning a vector the same length as the input
```{r}
map_dbl(flights, ~ sum(is.na(.x)))
```

# Part III: arrange

## `arrange`

### `arrange` reorder the rows, `-` means desc
### It takes a data frame and a set of column names to order by, if provide more than one column name, each additional column will be used to break ties in the values of preceding columns
### For example: reorder the flights by year, month and day
```{r, eval=FALSE}
arrange(flights,year,month,day)
```

## `arrange`
```{r, echo = FALSE}
arrange(flights,year,month,day)
```
## Missing values
### Missing values are always sorted at the end
```{r}
df<- tibble(x=c(5,2,NA))
df
arrange(df,x)
arrange(df,desc(x))
```

## Exercise 2
###1. How could we use `arrange` to sort all missing values to the start?
```{r}
df[order(is.na(df$x),decreasing= TRUE),]

arrange(df, desc(is.na(x)))
```

# Part IV: select

## `select`

### `select` pick variables by their names
### For example: select columns by name
```{r}
select(flights, year, month, day) %>%
  head(5)
#or
select(flights, year:day) %>%
  head(5)
```

## Examples
### Select all columns except those from year and day(inclusive)
```{r}
select(flights, -(year:day)) %>%
  head(10)
```

## Examples
### Select all columns contains "time"
```{r}
select(flights,contains("time")) %>%
  head(10)
```

## `rename`
### rename variables
```{r}
rename(flights, tail_num = tailnum)
```

## `everything`
### Use it to move a handful of variables to the start of the data frame
```{r}
select(flights, flight, everything())
```

# Part VI: summarise

## `summarise` or `summarize`

### `select` collapses a data frame to a single row
### For example: find the mean of `dep_delay`
```{r}
#na.rm = TRUE can excluding missing values from analysis
summarize(flights, delay=mean(dep_delay, na.rm=TRUE))
summarize(flights, delay=mean(dep_delay))
#group_by(summarize() is not terribly useful unless we pair it with group_by())
by_day <- group_by(flights, year, month, day)
summarise(by_day, delay=mean(dep_delay, na.rm=TRUE))
```

## `count`
### count the dest
```{r}
not_cancelled<-flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))
not_cancelled %>%
  count(dest)
```

## `count`
###count the total number of miles a plane flew
```{r}
not_cancelled %>%
  count(tailnum, wt = distance)
```

## Exercise 3
### Q1.Find the average of `dep_delay` for each dates
```{r}
flights %>%
  group_by(year, month, day) %>%
  summarize(mean = mean(dep_delay, na.rm = TRUE))
```

## Exercise 3
### Q2. Why is distance to some destinations more variable than to others?
```{r}
not_cancelled %>%
  group_by(dest)%>%
  summarise(distance_sd=sd(distance))%>%
  arrange(desc(distance_sd))
```

## Exercise 3
### Q3. What proportion of flights are delayed by more than one hour?
```{r}
not_cancelled %>%
  group_by(year, month, day)%>%
  summarise(hour_perc = mean(arr_delay > 60))
```

## Exercise 3
### Q4. Plot the proportion of cancelled flights related to the average delay
```{r, eval = FALSE}
flights %>%
  group_by(day) %>%
  summarise(cancelled = mean(is.na(dep_delay)),
            mean_dep = mean(dep_delay, na.rm = T),
            mean_arr = mean(arr_delay, na.rm = T)) %>%
  ggplot(aes(y = cancelled)) +
  geom_point(aes(x = mean_dep), colour = "red") +
  geom_point(aes(x = mean_arr), colour = "blue") +
  labs(x = "Avg delay per day", y = "Cancelled flights p day")
```

## Exercise 3
```{r, echo = FALSE}
flights %>%
  group_by(day) %>%
  summarise(cancelled = mean(is.na(dep_delay)),
            mean_dep = mean(dep_delay, na.rm = T),
            mean_arr = mean(arr_delay, na.rm = T)) %>%
  ggplot(aes(y = cancelled)) +
  geom_point(aes(x = mean_dep), colour = "red") +
  geom_point(aes(x = mean_arr), colour = "blue") +
  labs(x = "Avg delay per day", y = "Cancelled flights p day")
```

## Exercise 3
### Q5.Explore the relationship between the distance and average delay for each location
```{r, eval = FALSE}
by_dest <- group_by(flights, dest)
delay<- summarise(by_dest,
                  count=n(),
                  dist=mean(distance,na.rm=TRUE),
                  delay=mean(arr_delay,na.rm=TRUE)
)
delay<-filter(delay, count>20, dest!="HNL")

ggplot(delay,mapping = aes(dist,delay))+
  geom_point(aes(size=count),alpha=1/3)+
  geom_smooth(se=FALSE)
```

## Exercise 3
```{r, echo = FALSE}
by_dest <- group_by(flights, dest)
delay<- summarise(by_dest,
                  count=n(),
                  dist=mean(distance,na.rm=TRUE),
                  delay=mean(arr_delay,na.rm=TRUE)
)
delay<-filter(delay, count>20, dest!="HNL")

ggplot(delay,mapping = aes(dist,delay))+
  geom_point(aes(size=count),alpha=1/3)+
  geom_smooth(se=FALSE)
```
# Part VII: Brainstorm

##Find the top 10 data

###Find the 10 most delayed flights using a `ranking` function.  
```{r}
flights %>%
  filter(min_rank(-(dep_delay)) %in% 1:10)
```

##Find the top 10 data

###Find the 10 most delayed flights using a `top_n` function
```{r}
flights %>%
  top_n(10, dep_delay)
```


##Find the top 10 data

###Find the 10 most delayed flights using a `head` function
```{r}
flights %>%
  arrange(-(dep_delay)) %>%
  head(10)
```

##Find the top 10 data

###Find the 10 most delayed flights using a `head` function
```{r}
flights %>%
  mutate(ID = seq_along(dep_delay)) %>%
  select(ID, everything()) %>%
  arrange(-(dep_delay)) %>% 
  head(10)
```

##Find the top 10 data

###Find the rank 10th delayed flights using a `filter` function
```{r}
#If you want to find the rank first or last, we can use min() & first()
#or max() & last()
flights %>%
 arrange(-(dep_delay)) %>%
  filter(1:n()==10)
```

## For both, use `range`
```{r}
flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay)) %>%
  group_by(year, month, day) %>% 
 mutate(r = min_rank(desc(dep_time))) %>%
  filter(r %in% range(r))
```
