---
title: "R For Data Science" 
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    toc_float: true
date: "`r Sys.Date()`"
geometry: left = .5in, right = .5in, top = 1in, bottom = .6in
urlcolor: blue
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(prompt = TRUE, comment = "", collapse = TRUE) 
options(width = 100)
```

Load `tidyverse` with
```{r, message = FALSE}
library(tidyverse)
```
## `arrange` 
The `arrange` function offers a convenient way to order a data frame by columns.
Here is a book example: 
```{r}
df <- tibble(x = c(5, 2, NA), y = runif(3))
arrange(df, x)
arrange(df, desc(x))
```
The data (`df`) was arranged by the column `x` in either a descending order or ascending order.
A similar but standard function is to use the `order` function. For example,
```{r}
df[order(df$x),]
df[order(df$x, decreasing = TRUE),]
```
The `arrange` approach is more intuitive but the `order` approach is slighly faster.
```{r}
library(microbenchmark)
microbenchmark(arrange(df, x), df[order(df$x),])
```
Sorting with `is.na()` will be based on 0's and 1's (`FALSE` has a value of 0 and `TRUE` has a value of .).
```{r}
arrange(df, desc(is.na(x)))
order(is.na(df$x), decreasing = TRUE)
```